---
title: "PulmonDB - COVID19."
subtitle: "Regulons Selection: Lung"
date: "`r Sys.Date()`"
author: "Monica Padilla"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: cosmo
    highlight: textmate
    df_print: paged # kable, tibble
params:
  celllines : 0
  organs : 1
  lung : 0
  inputdir : ../regulons-analysis/out/lung-bladesdel/
  outdir : ../regulons-analysis/out/lung-bladesdel/plots/
  TF_Family_DBD_hsapiens_db : ~/Documents/RegGenoLab/dbs/TF_Family_DBD-hsapiens.csv # cisBP tf family info
  metadata.fn : ~/Documents/RegGenoLab/COVID19/pyscenic/regulons-analysis/out/lung-bladesdel/blanco2020-desai2020-delorey2021_allLungSamples_ID-source-disease.csv # experiment metadata
  tfs.fn : tfs780-ordered.txt # TF as outputted from scenic loom, only through scenic counts filter
  auc_mtx.fn : auc_mtx.RData
  l2fc.fn : log2fc_aucmtx.csv
  RSS.fn : RSSs.csv # Regulons Specificity Scores
  ## From reticulateEnrich_.R
  # pvals from DA tests
  pval_mwu.fn : pvalues-mwu-2t.csv
  pval_klm.fn : pvalues-klm-2t.csv
  # adjusted pvals from DA tests
  adj_pvals_mwu.fn : pvalues-mwu-2t-adjusted.csv
  adj_pvals_klm.fn : pvalues-klm-2t-adjusted.csv
  # regulons/tfs selection from DA and/or enrichment
  tfs_DA_2t.fn : tfs_mwu_ks_fdr01-2t.RData
  tfs_DA_enrichLFC_2t.fn : tfs_DA01_enrichLFC_2t.RData
  tfs_targets.fn : tfs_targets_fltr.RData
---

```{r, eval=FALSE, echo=FALSE}
# global variables
# para probar el codigo sin hacer knit
params <- {}
# analysis
params$celllines <- 0
params$organs <- 0
params$lung <- 1
# dirs
params$inputdir <- "/home/user/Documents/RegGenoLab/COVID19/pyscenic/regulons-analysis/out/lung-bladesdel/"
params$outdir <- "/home/user/Documents/RegGenoLab/COVID19/pyscenic/regulons-analysis/out/lung-bladesdel/plots/"
# files
params$TF_Family_DBD_hsapiens_db <- "~/Documents/RegGenoLab/dbs/TF_Family_DBD-hsapiens.csv"
params$metadata.fn <- "~/Documents/RegGenoLab/COVID19/pyscenic/regulons-analysis/out/lung-bladesdel/blanco2020-desai2020-delorey2021_allLungSamples_ID-source-disease.csv"
# direct out
params$tfs.fn <- "tfs780-ordered.txt" # tfs guiding regs- after filtering
params$tfs_targets.fn <- "tfs_targets_fltr.RData"
params$RSS.fn <- "RSSs.csv"
params$auc_mtx.fn <- "auc_mtx.RData"
params$l2fc.fn <- "log2fc_aucmtx.csv"
# selected regs based on tests
params$tfs_DA_enrichLFC_2t.fn <- "tfs_DA01_enrichLFC_2t.RData"
params$tfs_DA_2t.fn <- "tfs_mwu_ks_fdr01-2t.RData"
# tests pvalues
params$adj_pvals_mwu.fn <- "pvalues-mwu-2t-adjusted.csv"
params$adj_pvals_klm.fn <- "pvalues-klm-2t-adjusted.csv"
params$pval_mwu.fn <- "pvalues-mwu-2t.csv"
params$pval_klm.fn <- "pvalues-klm-2t.csv"
#
params$sub_cases_rss_plots <- 1
# params$up = 1
```

```{r}
## Load functions
source("~/Documents/RegGenoLab/COVID19/pyscenic/regulons-selection-scripts/regs-analysis.R")
```



\

## **Selection of Regulons based on Differentially Activation and Specificity **

Use of Regulon Specificity Scores (also referred here as enrichment although it is not exactly) + Differential Activation (DA) of Regulons Analysis. This is an exploratory report on the parameters used to select regulons.
Main code for Regulons Selection = `reticulateEnrich.R`

```{r, include=FALSE}
# Dont run code in this document
knitr::opts_chunk$set(
  #eval = FALSE
  rows.print=11,
  cols.min.print=12
)
```

\

### **DA tests pvalues distributions** {.tabset .tabset-pills}

#### **Mann Whitney U Test**

_P-values_ distributions resulting from `Mann Whitney U Test` for Differential Activation of Regulons (measured from AUC scores per regulon, Regulon Activity Matrix). Non-adjusted _p-values_ are displayed.

Run `reticulateEnrich_.R` first.


```{r, message=FALSE}
library(tidyverse) ## gplot, dplyr, stringr
library(ggpubr) ## ggarange

pvalues <- as.data.frame(read_csv(paste0(params$inputdir,params$pval_mwu.fn), col_names = T))

plots <- vector('list', length(pvalues)) # list of plots
# create plots iteratively
for (i in 1:ncol(pvalues)) {
  exp <- colnames(pvalues)[i]
  print(exp)

  plots[[i]] <- ggplot(pvalues,aes(x=.data[[exp]])) + #<<
    geom_bar(fill="#9966bb") + # geom hist didnt work
    scale_x_binned() +
    labs(title = exp, x="pvalue", y="")
}
names(plots) <- names(pvalues)

# display plots together
p <- ggarrange(plots[[1]],ncol = 1, nrow = 1)
# add title
annotate_figure(p,
                top = text_grob("P-values Mann Whitney U Tests",face = "bold", size = 14))

png(paste0(params$outdir,"hist_pvalues-mwu.png"), width = 600, height = 600)
# add title
annotate_figure(p,
                top = text_grob("P-values Mann Whitney U Tests",face = "bold", size = 14))
dev.off()

# minimum p values per experiment
apply(pvalues,2,min)
```


Each pvalue corresponds to a TF/regulon being differentially active in a given experiment vs its control (not shown).

* pvalues from **TRUE** null hypothesis (corresponding to Error Type II) distribution --> **uniform distribution** in range [0,1]

* pvalues from **FALSE** null hypothesis distribution --> asymetric, skewed to zero


Most of these distributions show that pvalues are more skewed to zero than what we would expect from a uniform distribution thus indicating that most of them correspond to **FALSE** null hypothesis.

Using a too low significance level would lead to rejecting true positives.[1]


#### **Kolmogrov Test**

_P-values_ distributions resulting from `Kolmogrov Test` for Differential Activation of Regulons (measured from AUC scores per regulon, Regulon Activity Matrix). Non-adjusted _p-values_ are displayed.

```{r, message=FALSE}
library(tidyverse) ## gplot, dplyr, stringr
library(ggpubr) ## ggarange


pvalues2 <- as.data.frame(read_csv(paste0(params$inputdir,params$pval_klm.fn), col_names = T))
# rename cols for display in graphs
names(pvalues2) <- names(pvalues)

plots2 <- vector('list', length(pvalues2)) # list of plots
# create plots iteratively
for (i in 1:ncol(pvalues2)) {
  exp <- colnames(pvalues2)[i]
  print(exp)

  plots2[[i]] <- ggplot(pvalues2,aes(x=.data[[exp]])) + #<<
    geom_bar(fill="#9966bb") + # geom hist didnt work
    scale_x_binned() +
    labs(title = exp, x="pvalue", y="")

}
names(plots2) <- names(pvalues2)

# display plots together
p <- ggarrange(plots[[1]], ncol = 1, nrow = 1)
# add title
annotate_figure(p,
                top = text_grob("P-values Kolmogrov Tests",face = "bold", size = 14))

png(paste0(params$outdir,"hist_pvalues-klm.png"), width = 600, height = 600)
annotate_figure(p,
                top = text_grob("P-values Kolmogrov Tests",face = "bold", size = 14))
dev.off()

apply(pvalues,2,min)
```


### **DA tests adjusted pvalues distributions** {.tabset .tabset-pills}

#### **Mann Whitney U Test**

Get dataframes of adjusted pvalues from both test to work with locally:

```{r, eval=FALSE}
# on the cluster
# Run `reticulateEnrich_.R` first
library(tidyverse)

# retrieve adjusted pvalues from difregs script
pvalues_adj <- py$pvalues_adjusted
pvalues_adj2 <- py$pvalues_adjusted2


# turn to dataframes
pv.adj.mwu <- t(data.frame(matrix(unlist(pvalues_adj), nrow=length(pvalues_adj), byrow=TRUE)))
colnames(pv.adj.mwu) <- keys_ct
pv.adj.klm <- t(data.frame(matrix(unlist(pvalues_adj2), nrow=length(pvalues_adj2), byrow=TRUE)))
colnames(pv.adj.klm) <- keys_ct
write.csv(pv.adj.mwu, file = paste0(outdir,"pvalues-mwu-2t-adjusted.csv"), quote = F, row.names = F)
write.csv(pv.adj.mwu, file = paste0(outdir,"pvalues-klm-2t-adjusted.csv"), quote = F, row.names = F)

```

Make plots:

```{r}
library(tidyverse) ## gplot, dplyr, stringr
library(ggpubr) ## ggarange

pv.adj.mwu <- as.data.frame(
  read_csv(file = paste0(params$inputdir,params$adj_pvals_mwu.fn),col_names = T))
pvalues <- pv.adj.mwu # just changed this #<<

plots <- vector('list', length(pvalues)) # list of plots
# create plots iteratively
for (i in 1:ncol(pvalues)) {
  exp <- colnames(pvalues)[i]
  print(exp)

  plots[[i]] <- ggplot(pvalues,aes(x=.data[[exp]])) + #<<
    geom_bar(fill="#9966bb") + # geom hist didnt work
    scale_x_binned() +
    labs(title = exp, x="pvalue", y="")

}
names(plots) <- names(pvalues)


# display plots together
p <- ggarrange(plots[[1]], ncol = 1, nrow = 1)
annotate_figure(p,top = text_grob("Adjusted P-values Mann Whitney U Tests",face = "bold", size = 14)) # and this

png(paste0(params$outdir,"hist_adj-pvalues-mwu.png"), width = 600, height = 600)
annotate_figure(p,top = text_grob("Adjusted P-values Mann Whitney U Tests",face = "bold", size = 14)) # and this
dev.off()

# minimum p values per experiment
apply(pvalues,2,min)
```


#### **Kolmogrov Test**

```{r, message=FALSE}
library(tidyverse) ## gplot, dplyr, stringr
library(ggpubr) ## ggarange

pv.adj.klm <- as.data.frame(read_csv(paste0(params$inputdir,params$adj_pvals_klm.fn), col_names = T))
pvalues2 <- pv.adj.klm # just changed this #<<

# rename cols for display in graphs
names(pvalues2) <- names(pvalues)

plots2 <- vector('list', length(pvalues2)) # list of plots
# create plots iteratively
for (i in 1:ncol(pvalues2)) {
  exp <- colnames(pvalues2)[i]
  print(exp)

  plots2[[i]] <- ggplot(pvalues2,aes(x=.data[[exp]])) + #<<
    geom_bar(fill="#9966bb") + # geom hist didnt work
    scale_x_binned() +
    labs(title = exp, x="pvalue", y="")

}
names(plots2) <- names(pvalues2)

# display plots together
p <- ggarrange(plots[[1]], ncol = 1, nrow = 1)
# add title
annotate_figure(p,
                top = text_grob("Adjusted P-values Kolmogrov Tests",
                                face = "bold", size = 14)) # and this

png(paste0(params$outdir,"hist_adj-pvalues-klm.png"), width = 600, height = 600)
annotate_figure(p,
                top = text_grob("Adjusted P-values Kolmogrov Tests",
                                face = "bold", size = 14)) # and this
dev.off()

apply(pvalues,2,min)
```


### **FDR threshold on adjusted p-values** {.tabset .tabset-pills}

#### **Mann Whitney U Test**

Import dataset and call source script with functions:

```{r, message=FALSE}
library(tidyverse)

## call functios
source("~/Documents/RegGenoLab/COVID19/PulmonDB_COVID/COVID/bulk/rScripts/FDRthreshold_functions.R")

## dataset
pv.adj.mwu <- as.data.frame(read_csv(paste0(params$inputdir,params$adj_pvals_mwu.fn), col_names = T))
pv.adj.mwu <- as.data.frame(pv.adj.mwu)
# rename cols for display in graph
#head(pv.adj.mwu)
```


```{r, message=FALSE}
thrs.mwu <- regsPerFDRthr(pv.adj.mwu, fdr_lowth = 0.01)
thrs.mwu
```

Se buscó que el `FDR threshold` (entre `0.2` y `0.1`) elegido permitiera que nos quedaramos con alrededor de `100 regulones` por experimento. Esto puesto que el set de regulones se someterá a otros filtros:
- pvalor ajustado en test de Kolmogrov
- LFC de enriquecimiento de regulon en el experimento dado
En conjunto, estos filtros darán lugar a un set de regulones cuya especificidad y relevancia en el experimento dado sean relativamente confiables.

El `FDR threshold` elegido por experimento se especifica aquí:

```{r, message=FALSE}
fdrs.mwu <- getChosenFDRth(thrs.mwu, min.regs = c(100,30))
fdrs.mwu
```

#### **Kolmogrov Test**


```{r, message=FALSE}
thrs.klm <- regsPerFDRthr(pv.adj.klm, fdr_lowth = 0.01)
thrs.klm
```


El `FDR threshold` elegido por experimento se especifica aquí:

```{r, message=FALSE}
fdrs.klm <- getChosenFDRth(thrs.klm, min.regs = c(100,30))
fdrs.klm
```


### **Regulons Counts After Filtering**

\

#### **DA:**

After taking common Differentialy Activated Regulons from both tests:

```{r}
tfs_da.fn <- paste0(params$inputdir, params$tfs_DA_2t.fn)
tfs_da <- readRDS(tfs_da.fn)
# at reticulateEnrich line 117
## check how many regulons were kept
t(as.data.frame(lapply(tfs_da,length)))
```

✅ Filter Applied

\

Differentiate between **Up** and **Down Regulated** regulons based on logFC (infection/mock):

```{r}
library(tidyverse)

## Load Data
# auc_mtx
auc_mtx <- readRDS(paste0(params$inputdir,params$auc_mtx))

# Load data: TFs names (in order after just scenic counts cutoff)
tfs <- as.data.frame(read_csv(paste0(params$inputdir,params$tfs.fn), col_names = F))[,1]
tfs <- stringr::str_remove(tfs, "_\\(\\+\\)")
tfs <- tfs[match(str_remove(colnames(auc_mtx),"_\\(\\+\\)"), tfs)]

# Load and tidy data: log 2 FC (auc_mtx)
l2fc <- as.data.frame(read_csv(paste0(params$inputdir,params$l2fc.fn), col_names = T))
rownames(l2fc) <- colnames(auc_mtx)
head(l2fc)

## match dfs rows (tfs)
#l2fc <- l2fc[match(str_replace(tfs,"$","_\\(\\+\\)"),rownames(l2fc)),]

## Make lists of up/down regulons
tfs_da_up = list()
tfs_da_down = list()
for (exp in names(tfs_da)) {
  tfs_da_up[[exp]] <- tfs_da[[exp]][tfs_da[[exp]] %in%  rownames(l2fc)[which(l2fc[exp] > 0)]]
  tfs_da_down[[exp]] <- tfs_da[[exp]][tfs_da[[exp]] %in% rownames(l2fc)[which(l2fc[exp] < 0)]]
}

# make df of counts of regulons
n_up <- t(as.data.frame(lapply(tfs_da_up,length)))[,1]
n_down <- t(as.data.frame(lapply(tfs_da_down,length)))[,1]
data.frame(up = n_up, down = n_down)
```

✅ Filter Applied  for  **Upregulated** Regulons

***

#### **DA + specificity:**

After filtering regulons for those that had an optimal `LFC enrichment` :

```{r}
tfs_da_enrich.fn <- paste0(params$inputdir, params$tfs_DA_enrichLFC_2t.fn)
tfs_da_enrich <- readRDS(tfs_da_enrich.fn)
names(tfs_da_enrich) <- names(tfs_da)
# at reticulateEnrich line 117
## check how many regulons were kept
t(as.data.frame(lapply(tfs_da_enrich,length)))
```

\

Differentiate between **Up** and **Down Regulated** regulons again:

```{r}
## Make lists of up/down regulons
tfs_da_enrich_up = list()
tfs_da_enrich_down = list()
for (exp in names(tfs_da_enrich)) {
  tfs_da_enrich_up[[exp]] <-
    tfs_da_enrich[[exp]][tfs_da_enrich[[exp]] %in% rownames(l2fc)[which(l2fc[exp] > 0)]]
  tfs_da_enrich_down[[exp]] <-
    tfs_da_enrich[[exp]][tfs_da_enrich[[exp]] %in% rownames(l2fc)[which(l2fc[exp] < 0)]]
}

# make df of counts of regulons
n_up <- t(as.data.frame(lapply(tfs_da_enrich_up,length)))[,1]
n_down <- t(as.data.frame(lapply(tfs_da_enrich_down,length)))[,1]
data.frame(up = n_up, down = n_down)
```

For regulons that were `Upregulated` and enriched, this filter keeps a decent number of regulons, but that's not the case for `Downregulated` regulons.

✅ Filter Applied  for  **Upregulated** Regulons


<!-- ### **FDR threshold less strict**  -->
<!-- ### **Regulons Counts After Filtering 2** -->


### **Volcano Plots**

List of all Differentially Activated Regulons:

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(ggrepel)

############################################
## Plots 2t - 2 tails test

# Load and tidy data: regulons/tfs that were in DA + enriched
paste0("using set of DA regulons (tfs): ",tfs_da.fn)
tfs_da.fn <- paste0(params$inputdir, params$tfs_DA_2t.fn)
tfs_da <- readRDS(tfs_da.fn)
dfs <- list2Dfs(tfs_da)
dfs$Lung

# Load and tidy data: adjusted pvalues, format
paste0("using pvalues from test : ", params$adj_pvals_mwu.fn)
pvalues <- as.data.frame(read_csv(paste0(params$inputdir,params$adj_pvals_mwu.fn), col_names = T))
rownames(pvalues) <- tfs

## Graph a Volcano Plot per condition
plots <- volcanos_per_exp(pvalues = pvalues, tfs_da = tfs_da,
                          test = "2t", save_png=T, outdir = params$outdir)

plots

############################################
## Make lists of up/down regulons
tfs_da_up = list()
tfs_da_down = list()
for (exp in names(tfs_da)) {
  tfs_da_up[[exp]] <- tfs_da[[exp]][tfs_da[[exp]] %in%  rownames(l2fc[exp] > 0)]
  tfs_da_down[[exp]] <- tfs_da[[exp]][tfs_da[[exp]] %in% rownames(l2fc[exp] > 0)]
}
```


List of all Differentially Activated **Upregulated** Regulons:

```{r, message=FALSE, warning=FALSE}
dfs <- list2Dfs(tfs_da_up)
dfs$Lung
```

List of all Differentially Activated **Downregulated** Regulons:
```{r, message=FALSE, warning=FALSE}
dfs <- list2Dfs(tfs_da_down)
dfs$Lung
```



### **References**


[1] https://www.cienciadedatos.net/documentos/19b_comparaciones_multiples_correccion_p-value_fdr

[2] https://yulab-smu.top/biomedical-knowledge-mining-book/clusterprofiler-go.html

```{r}
sessionInfo()
```
